A Gentle Introduction into Structural Causal Models

### 2021-05-17 draft.Rmd
hardt chapter

Most students have heard the phrase 'correlation does not imply causation.'
While correlation implies co-occurrence, for many problems correlation is not enough. 
Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
For many problems we want to understand causal relationships between variables.
There are different approaches on how to model causal relationships.
This paper focuses on Structural Causal Models and Bayesian Causal Networks.
Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
SCMs are non-parametric structural equations models with added features.
Existing literature has provided an excellent introduction to SCMs and BCNs. 
@pearl2009causality provided a comprehensive introduction into this topic with his book 'Causality'.
His work addressed misconceptions in social sciences, causality and statistics.
His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
@peters_elements_2017 address the relationship of causality, physical sciences, and causal inference.
Their work highlights the importance of time in causality and inference based on observational data alone. 
E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
@tarka2018overview discuss the history of SEMs and SCMs.
The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
I structure the rest of the paper as follows: 
Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different causal differencecs, fundamental differences and a brief history of the methods. 
Section 3 addresses Pearls Causal Hierachy

@@

@@

The circle nodes represent the observed variables.
The arrows are our structural equations and depict the relationship between variables.
^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks. For further information see Pearl (2009). Due to the confined scope of this paper, I will not explore cyclic structures.]

@@

To contextualise causal methods, Judea Pearl (2009) introduced the hierarchy of causation.
Pearl focuses on three layers: association, intervention and counterfactuals (Table 1).

@@

A higher level implies more detailed knowledge of the relationship between the variables.
The first query is association, where we examine relationships based on observations.

@@

## Prediction

>>>>

Most students have heard the phrase 'correlation does not imply causation.'
While correlation implies co-occurrence, for many problems correlation is not enough. 
Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
For many problems we want to understand causal relationships between variables.
There are different approaches on how to model causal relationships.
This paper focuses on Structural Causal Models and Bayesian Causal Networks.
Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
Summarized, *'they are simulators that generate a distribution based noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
SCMs are non-parametric structural equations models with added features.
Existing literature has provided an excellent introduction to SCMs and BCNs. 
@pearl2009causality provided a comprehensive introduction into this topic with his book 'Causality'.
His work addressed misconceptions in social sciences, causality and statistics.
His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
@peters_elements_2017 address the relationship of causality, physical sciences, and causal inference.
Their work highlights the importance of time in causality and inference based on observational data alone. 
E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
@tarka2018overview discuss the history of SEMs and SCMs.
The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
I structure the rest of the paper as follows: 
Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different causal differencecs, fundamental differences and a brief history of the methods. 
Section 3 addresses Pearls Causal Hierachy

@@

**Established Conditions:**

Definition: **SUTVA**

'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]

Definition: **Consistency**

The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]

Definition: **Ignorability**

The potential outcomes are conditionally independent of treatment given some set of deconfounding variables. [@hardtrecht]
(perfect RCT)

- First tow hold for SCM counterfactuals 
- third not testable but can check via backdoor criterion in SCM

@@

The circle nodes represent the observed variables.
The arrows are our structural equations and depict the relationship between variables.
^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks. For further information see  @pearl2009causality. Due to the confined scope of this paper, I will not explore cyclic structures.

Potential outcomes treat outcomes merely as potential outcomes rather rather than true outcomes. 
- Only observe outcomes for units that are treated
- ATE is a typical query
- Every SCM can generate PO 
- Cannot derive SCM from PO alone 
- SCM has more concepts : causal graphs, mediating paths, counterfactuals for every variable etc. 
- PO Easier to apply because apply to observational data [@hardtrecht]

@@

To contextualise causal methods,  @pearl2009causality introduced the hierarchy of causation.
Pearl focuses on three layers: association, intervention and counterfactuals (Table 1).

@@

A higher level implies more detailed knowledge of the relationship between the variables.
 @pearl2009causality introduced the do-calculus.
 The do-calculus is a mathematical framework allowing for active substitution (E.g. $X:=10$) [@hardtrecht].
The first query is association, where we examine relationships based on observations.

@@

## Association
